### 2024-03-19 manuscript.Rmd
Continue working on the project

@luhmannSubjectiveWellbeingAdaptation2012 has done a meta-analysis on adaptation to life events focusing on SWB. Evaluating the robustness of the studies focusing on SWB and adaptation is difficult, since it is impossible to manipulate life events experimentally and cross-sectional studies have a lack of internal validity [@luhmannSubjectiveWellbeingAdaptation2012]. All of the following statements are based on the meta-analysis by @luhmannSubjectiveWellbeingAdaptation2012.

@@

Studies focusing on the hedonic treadmill can not be adequatly characterized by single effect sizes, since the temporal development of SWB is of interest. So @luhmannSubjectiveWellbeingAdaptation2012 used a method that takes time into account. For each life event, they fitted a logarithmic model using the effect sizes reported in original studies. An adaptation is characterized by a change parameter, which is in the opposite direction of the initial reaction. For example, having a positive reaction (higher SWB) after a life event, then a negative change in SWB, indicates that an adaptation is taking place.

@@

@@

@@

![](material/setpoint_definitions_swb.drawio.png)

- SWBSetPoint: SWB people return to
- AdaptationRate: How fast the adaptation happens
  - AdaptationRatePerson: Personenvariable
- InitialReaction: Change in SWB directly after / at the time of the LifeEvent
- LifeEvent: Status change

## Model

# Implementation

We made small changes to the theoretical model, in order to make it suitable for an implementation. We are making the following assumptions:

- We do not take anticipatory effects into account. SWB tends to change before major life events, if people know that something is going to happen. Before marriage as life event people can already anticipate their wedding, which leads to an increase in SWB.
- AWB and CWB are summarized into a single SWB construct.
- Life Event and Initial Reaction are reduced to Initial Reaction (appraisal of the situation), since only the latter is important for the following psychological processes.

*VAST display of the model here*

@@

# Simulation
```{r, include=FALSE}
source("simulation/simulation.R")
```

@@

>>>>

@luhmannSubjectiveWellbeingAdaptation2012 has done a meta-analysis on adaptation to life events focusing on SWB, they included 313 studies focusing on 8 different major life events. The total number of participants was 65911. Evaluating the robustness of the studies focusing on SWB and adaptation is difficult, since it is impossible to manipulate life events experimentally and cross-sectional studies have a lack of internal validity [@luhmannSubjectiveWellbeingAdaptation2012]. All of the following statements are based on the meta-analysis by @luhmannSubjectiveWellbeingAdaptation2012.

@@

Studies focusing on the hedonic treadmill can not be adequately characterized by single effect sizes, since the temporal development of SWB is of interest. So @luhmannSubjectiveWellbeingAdaptation2012 used a method that takes time into account. For each life event, they fitted a logarithmic model using the effect sizes reported in original studies. An adaptation is characterized by a change parameter, which is in the opposite direction of the initial reaction. For example, having a positive reaction (higher SWB) after a life event, then a negative change in SWB, indicates that an adaptation is taking place.

@@

The visual representations of construct definitions and their relationships with each other according to the theory are done using VAST [@leisingVisualArgumentStructure2023].

@@

The central constructs of the *Set Point Theory* are SWB, set point of SWB, life event, initial reaction and adaptation rate. We added the constructs emotional reactivity, and personal adaptation rate. The VAST displays for the construct definitions are shown below.

@@

![](material/setpoint_definitions-SWB.drawio.png)

### Other relevant constructs

![](material/setpoint_definitions-Others.drawio.png)

We made small changes to the theoretical model, in order to make it suitable for an implementation.

@luhmannSubjectiveWellbeingAdaptation2012 clearly state that the adaptation after negative life events is slower. This would complicate the model and it is out of the scope of this project.

We also do not take anticipatory effects into account. SWB tends to change before major life events, if people know that something is going to happen. For example, before their marriage people can already anticipate their wedding, which leads to an increase in SWB. This is very well reflected in the empirical findings but it is not a part of the *Set Point Theory*.

Empirical findings show that AWB and CWB are affected separately by life events, their adaptations are also independent. @dienerHedonicTreadmillRevising2006a also have this in their models, as extension (3) but they do not state how separate well-being variables are affected by life events, how different their set points are, and what they change in the adaptation process.

![](material/setpoint_definitions-Theory.drawio.png)

@@

# Simulation and Evaluation of the Model
```{r, include=FALSE}
source("simulation/simulation.R")
```

@@

# (APPENDIX) Appendix {-}

# Statements

```{r child="./material/statements.Rmd"}
```

### 2024-03-19 material/statements.Rmd
Continue working on the project

| Identifier | Citation                                                                                                                                                                                                                                                                                                                                                                                                                                                                                               | Reference                                                            | Relationship?                      | Comments?                                                                      |     |
| ---------- | ------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------ | -------------------------------------------------------------------- | ---------------------------------- | ------------------------------------------------------------------------------ | --- |
| A          | A notable finding was that the initial reaction and the rate of adaptation varied considerably between different life events.                                                                                                                                                                                                                                                                                                                                                                          | Luhmann et al. (2012)                                                | Appraisal  - c → Adaptation Rate   | Valence of the life event changes the adaptation rate.                         |     |
| A1         | For instance, the initial reaction to marriage was positive, but subsequent adaptation was fast and completed after 2 years on average (Lucas et al., 2003).                                                                                                                                                                                                                                                                                                                                           |                                                                      |                                    | positive valence: fast adaptation                                              |     |
| A2         | In contrast, the rate of adaptation was much slower for negative events such as the onset of disability (Lucas, 2007b), widowhood (Lucas et al., 2003; Specht, Egloff, & Schmukle, 2011), and divorce (Lucas, 2005).                                                                                                                                                                                                                                                                                   |                                                                      |                                    | negative valence: slow adaptation                                              |     |
| A3         | The findings were perhaps most striking for unemployment: After the beginning of unemployment, mean levels of SWB were significantly below baseline, even if people became reemployed (Lucas et al., 2004). Repeated unemployment spells aggravate this effect (Luhmann & Eid, 2009).                                                                                                                                                                                                                  |                                                                      |                                    |                                                                                |     |
| B          | Together, these studies indicate that major life events can have strong effects on SWB and that the strength of these effects varies, depending on the life events considered.                                                                                                                                                                                                                                                                                                                         |                                                                      |                                    |                                                                                |     |
| B1         | major life events can have strong effects on SWB                                                                                                                                                                                                                                                                                                                                                                                                                                                       |                                                                      | Life Event -c→ SWB                 |                                                                                |     |
| B2         | the strength of these effects varies, depending on the life events considered                                                                                                                                                                                                                                                                                                                                                                                                                          |                                                                      | Life Event -c→ intensity of change |                                                                                |     |
| C          | SWB relates to how people feel and think about their lives.                                                                                                                                                                                                                                                                                                                                                                                                                                            | Indirectly from Diener (1984) via Luhmann et al. (2012)              | n                                  |                                                                                |     |
| D          | AWB and CWB are distinct constructs                                                                                                                                                                                                                                                                                                                                                                                                                                                                    | Indirectly from Lucas, Diener & Suh (1996) vie Luhmann et al. (2012) |                                    |                                                                                |     |
| D1         | Affective well-being (AWB) refers to the presence of pleasant affect (e.g., feelings of happiness) and the absence of unpleasant affect (e.g., depressed mood).                                                                                                                                                                                                                                                                                                                                        |                                                                      | n                                  |                                                                                |     |
| D2         | Cognitive well-being (CWB) refers to the cognitive evaluation of life overall (i.e., global life satisfaction) as well as of specific life domains (e.g., job satisfaction or marital satisfaction).                                                                                                                                                                                                                                                                                                   |                                                                      | n                                  |                                                                                |     |
| E          | It is therefore plausible that other external life circumstances, such as life events, have differential effects on AWB and CWB and that adaptation of AWB and CWB does not occur at the same rate.                                                                                                                                                                                                                                                                                                    | Luhmann et al. (2012)                                                |                                    | Having AWB and CWB separated in the model could be a good idea                 |     |
| F          | Life events have been examined from two major perspectives (Filipp & Aymanns, 2009): a stress perspective and a developmental perspective.                                                                                                                                                                                                                                                                                                                                                             | Luhmann et al. (2012)                                                |                                    |                                                                                |     |
| F1         | From the stress perspective, life events are viewed as specific types of stressors (e.g., Park, 2010; Segerstrom & Miller, 2004). Stressful life events are all events that significantly disturb the daily routine (Turner & Wheaton, 1997). This definition explicitly includes desirable events such as marriage or vacation (Holmes & Rahe, 1967).                                                                                                                                                 |                                                                      | Life Event -c→ Stress              | The stress perspective is easier to model.                                     |     |
| F2         | By contrast, minor stressors such as daily hassles and uplifts (Kanner, Coyne, Schaefer, & Lazarus, 1981) are distinct from stressful life events. From the developmental perspective, life events are viewed as specific transitions. Transitions are defined as a “discontinuity in a person’s life space of which he is aware and which requires new behavioural responses” (Hopson & Adams, 1976, p. 24).                                                                                          |                                                                      |                                    |                                                                                |     |
| F3         | For the purpose of this metaanalysis, we developed a narrower definition of life events that integrates elements from both perspectives. According to our working definition, life events are time-discrete transitions that mark the beginning or the end of a specific status. A status is a nominal variable with at least two levels.                                                                                                                                                              | Luhmann et al. (2012)                                                | Life Event -c→ Status              |                                                                                |     |
| G          | the process perspective explicitly predicts the trajectory of SWB over time within individuals                                                                                                                                                                                                                                                                                                                                                                                                         | Luhmann et al. (2012)                                                |                                    |                                                                                |     |
| G1         | The adaptation process is initiated by an external stimulus (e.g., a major life event) that causes a physiological or psychological response (e.g., decreased SWB). Over time, the responsiveness diminishes and the level of SWB returns to its preevent level.                                                                                                                                                                                                                                       | Luhmann et al. (2012)                                                |                                    | IMPORTANT<br><br>This description can be used to model the adaptation process! |     |
| H          | Most major life events are—at least to some extent— controllable and predictable and can therefore be anticipated. This anticipation might cause a specific hedonic reaction even before the event occurred: If the spouse is terminally ill, the hedonic reaction to bereavement starts long before the spouse actually dies. These so-called anticipatory or lead effects can be observed months or even several years before the occurrence of the event (Clark, Diener, Georgellis, & Lucas, 2008) | Luhmann et al. (2012)                                                |                                    |                                                                                |     |

>>>>

- ”MA” (meta-analysis): Luhmann et al. (2012)
- “TP” (theory paper): Diener et al. (2006)
- “OI” (original idea): Ideas / assumptions I added myself.
- “OI(MA/TP/MA+TP)”: Original ideas I extracted from things mentioned  in the meta-analysis or theory paper. It is also possible to combine insights from both papers.

| ID  | Citation                                                                                                                                                                                                                                                                                                                                                                                                                                                                                               | Reference                                    |
| --- | ------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------ | -------------------------------------------- |
| A   | A notable finding was that the initial reaction and the rate of adaptation varied considerably between different life events.                                                                                                                                                                                                                                                                                                                                                                          | MA                                           |
| A1  | For instance, the initial reaction to marriage was positive, but subsequent adaptation was fast and completed after 2 years on average (Lucas et al., 2003).                                                                                                                                                                                                                                                                                                                                           | MA                                           |
| A2  | In contrast, the rate of adaptation was much slower for negative events such as the onset of disability (Lucas, 2007b), widowhood (Lucas et al., 2003; Specht, Egloff, & Schmukle, 2011), and divorce (Lucas, 2005).                                                                                                                                                                                                                                                                                   | MA                                           |
| A3  | The findings were perhaps most striking for unemployment: After the beginning of unemployment, mean levels of SWB were significantly below baseline, even if people became reemployed (Lucas et al., 2004). Repeated unemployment spells aggravate this effect (Luhmann & Eid, 2009).                                                                                                                                                                                                                  | MA                                           |
| B   | Together, these studies indicate that major life events can have strong effects on SWB and that the strength of these effects varies, depending on the life events considered.                                                                                                                                                                                                                                                                                                                         | MA                                           |
| B1  | major life events can have strong effects on SWB                                                                                                                                                                                                                                                                                                                                                                                                                                                       | MA                                           |
| B2  | the strength of these effects varies, depending on the life events considered                                                                                                                                                                                                                                                                                                                                                                                                                          | MA                                           |
| C   | SWB relates to how people feel and think about their lives.                                                                                                                                                                                                                                                                                                                                                                                                                                            | Indirectly Diener (1984) via MA              |
| D   | AWB and CWB are distinct constructs                                                                                                                                                                                                                                                                                                                                                                                                                                                                    | Indirectly Lucas, Diener & Suh (1996) via MA |
| D1  | Affective well-being (AWB) refers to the presence of pleasant affect (e.g., feelings of happiness) and the absence of unpleasant affect (e.g., depressed mood).                                                                                                                                                                                                                                                                                                                                        | MA                                           |
| D2  | Cognitive well-being (CWB) refers to the cognitive evaluation of life overall (i.e., global life satisfaction) as well as of specific life domains (e.g., job satisfaction or marital satisfaction).                                                                                                                                                                                                                                                                                                   | MA                                           |
| E   | It is therefore plausible that other external life circumstances, such as life events, have differential effects on AWB and CWB and that adaptation of AWB and CWB does not occur at the same rate.                                                                                                                                                                                                                                                                                                    | MA                                           |
| F   | Life events have been examined from two major perspectives (Filipp & Aymanns, 2009): a stress perspective and a developmental perspective.                                                                                                                                                                                                                                                                                                                                                             | MA                                           |
| F1  | From the stress perspective, life events are viewed as specific types of stressors (e.g., Park, 2010; Segerstrom & Miller, 2004). Stressful life events are all events that significantly disturb the daily routine (Turner & Wheaton, 1997). This definition explicitly includes desirable events such as marriage or vacation (Holmes & Rahe, 1967).                                                                                                                                                 | MA                                           |
| F2  | By contrast, minor stressors such as daily hassles and uplifts (Kanner, Coyne, Schaefer, & Lazarus, 1981) are distinct from stressful life events. From the developmental perspective, life events are viewed as specific transitions. Transitions are defined as a “discontinuity in a person’s life space of which he is aware and which requires new behavioural responses” (Hopson & Adams, 1976, p. 24).                                                                                          | MA                                           |
| F3  | For the purpose of this metaanalysis, we developed a narrower definition of life events that integrates elements from both perspectives. According to our working definition, life events are time-discrete transitions that mark the beginning or the end of a specific status.                                                                                                                                                                                                                       | MA                                           |
| F4  | A status is a nominal variable with at least two levels.                                                                                                                                                                                                                                                                                                                                                                                                                                               |                                              |
| G   | the process perspective explicitly predicts the trajectory of SWB over time within individuals                                                                                                                                                                                                                                                                                                                                                                                                         | MA                                           |
| G1  | The adaptation process is initiated by an external stimulus (e.g., a major life event) that causes a physiological or psychological response (e.g., decreased SWB). Over time, the responsiveness diminishes and the level of SWB returns to its preevent level.                                                                                                                                                                                                                                       | MA                                           |
| H   | Most major life events are—at least to some extent— controllable and predictable and can therefore be anticipated. This anticipation might cause a specific hedonic reaction even before the event occurred: If the spouse is terminally ill, the hedonic reaction to bereavement starts long before the spouse actually dies. These so-called anticipatory or lead effects can be observed months or even several years before the occurrence of the event (Clark, Diener, Georgellis, & Lucas, 2008) | MA                                           |
| I   | The original treadmill theory suggested that people return to a neutral set point after an emotionally significant event. However, decades of research show that this part of the hedonic treadmill theory is wrong. Instead, most people are happy most of the time (Diener & Diener, 1996).                                                                                                                                                                                                          | TP                                           |
| J   | The empirical research that has been conducted since the publication of Brickman and Campbell (1971) reveals that if people do have set points, they vary considerably across individuals.                                                                                                                                                                                                                                                                                                             | TP                                           |
| K   | But we have found individual differences in the rate and extent of adaptation that occurs even to the same event. In our longitudinal studies, the size and even the direction of the change in life satisfaction varied considerably across individuals.                                                                                                                                                                                                                                              | TP                                           |
| L   | (not a direct citation) Set point of SWB = the level of SWB people return to after life events.                                                                                                                                                                                                                                                                                                                                                                                                        | TP                                           |
| M   | The difference between the current SWB and the set point affects the change rate. If the difference is larger, expected change is also larger.                                                                                                                                                                                                                                                                                                                                                         | OI(MA+TP)                                    |
| N   | (not a direct citation) People react differently to life events                                                                                                                                                                                                                                                                                                                                                                                                                                        | MA                                           |

### 2024-03-19 material/variable_definitions.Rmd
Continue working on the project

| **Construct** | **Scale Level** | **Range** | **Anchors** |
| --------- | ----------- | ----- | ------- |
| Subjective Well-Being | continuous | [0; 1] | 0 = absolutely not satisfied with life \newline 1 = maximal life satisfaction |
| Subjective Well-Being Set Point | continuous | [0; 1] | 0 = absolutely not satisfied with life \newline 1 = maximal life satisfaction |
| Appraisal of Life Event | continuous | [-1; 1] | -1 = negative \newline 0 = neutral \newline 1 = positive |

>>>>

| **Construct** | **Scale Level** | **Range** | **Anchors** |
| --------- | ----------- | ----- | ------- |
| SWB | continuous | [0; 1] | 0 = absolutely not satisfied with life \newline 1 = maximal life satisfaction |
| SWBSetPoint | continuous | [0; 1] | 0 = absolutely not satisfied with life \newline 1 = maximal life satisfaction |
| LifeEvent | continous | [-1; 1] | -1 = objectively very negative \newline 0 = objectively neutral \newline 1 = objectively very positive |
| EmotionalReactivity | continuous | [0; 1] | 0 = emotional state is not influenced by external events \newline 1 = external events influence emotional states very strongly |
| InitialReaction | continuous | [-1; 1] | -1 = negative \newline 0 = neutral \newline 1 = positive |
| AdaptationRate | continuous | [0; 1] | 0 = no adaptation \newline 1 = fastest possible adaptation |
| AdaptationRatePerson | continous | [0; 1] | 0 = lowest possible tendency for adaptation \newline 1 = highest possible tendency for adaptation |
